---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data
```{r, echo=TRUE}
# Loading and preprocessing the data
data.initial <- read.csv(unz("activity.zip", "activity.csv"), sep=",")
data.initial$date <- as.Date(data.initial$date)
```

## What is mean total number of steps taken per day?
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# What is mean total number of steps taken per day?
# ignore the missing values in the dataset
data <- data.initial[complete.cases(data.initial), ]
# 1. Calculate the total number of steps taken per day
library(dplyr)
data.per.day <- data %>% 
                  group_by(date) %>% 
                  summarise(sum_steps=sum(steps))

# 2. Make a histogram of the total number of steps taken each day
library(ggplot2)
histogram <- ggplot(data = data.per.day, aes(x =sum_steps)) + 
  geom_histogram(bins = 9, fill='skyblue2', color='skyblue4') + 
  stat_bin(bins = 9,geom="text", aes(label=..count..), vjust=-0.5) +
  ggtitle("Histogram of steps per days") +
  xlab("days") + ylab("Number of steps")
histogram

# 3. Calculate and report the mean and median of the total number of steps taken per day
mean.steps.per.days <- mean(data.per.day$sum_steps)
# print(paste("Mean of the total number of steps per day: ", mean.steps.per.days))
median.steps.per.days <- median(data.per.day$sum_steps)
# print(paste("Median of the total number of steps per day: ", median.steps.per.days))
```

Mean of the total number of steps per day: **`r format(mean.steps.per.days, digits=6)`**   
Median of the total number of steps per day: **`r median.steps.per.days`**

## What is the average daily activity pattern?
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# What is the average daily activity pattern?
# 1. Make a time series plot (i.e. type = "l"type = "l") of the 5-minute interval (x-axis) and 
# the average number of steps taken, averaged across all days (y-axis)
data.per.interval <- data %>% 
                      group_by(interval) %>% 
                      summarise(avg_steps=mean(steps))
plot(data.per.interval$interval, 
     data.per.interval$avg_steps, 
     type="l",
     main="Average number of steps per intervals",
     xlab="Interval",
     ylab="Average number of steps")

# 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
interval.max.steps <- data.per.interval$interval[data.per.interval$avg_steps == max(data.per.interval$avg_steps)]
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps? Interval **`r interval.max.steps`**


## Imputing missing values

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Imputing missing values
# 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
nb.missing.rows <- nrow(data.initial)-nrow(data)

# 2. Devise a strategy for filling in all of the missing values in the dataset. 
# The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, 
# or the mean for that 5-minute interval, etc.
# 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
data.filled.na <- data.initial %>% 
                  inner_join(data.per.interval, by= "interval") %>%
                  mutate(steps = coalesce(steps, avg_steps)) %>%
                  select(interval, date, steps)

# 4.Make a histogram of the total number of steps taken each day and Calculate and 
data.per.day.filled <- data.filled.na %>% 
                        group_by(date) %>% 
                        summarise(sum_steps=sum(steps))

histogram.filled <- ggplot(data = data.per.day.filled, aes(x =sum_steps)) + 
  geom_histogram(bins = 9, fill='skyblue2', color='skyblue4') + 
  stat_bin(bins = 9,geom="text", aes(label=..count..), vjust=-0.5) +
  ggtitle("Histogram of steps per days (NA filled)") +
  xlab("days") + ylab("Number of steps")
histogram.filled
# report the mean and median total number of steps taken per day.
mean.steps.per.days.filled <- mean(data.per.day.filled$sum_steps)
# print(paste("Mean of the total number of steps per day (NA filled): ", mean.steps.per.days.filled))
median.steps.per.days.filled <- median(data.per.day.filled$sum_steps)
# print(paste("Median of the total number of steps per day (NA filled): ", median.steps.per.days.filled))
# Do these values differ from the estimates from the first part of the assignment?
mean.steps.filled.difference <- mean.steps.per.days.filled - mean.steps.per.days
median.steps.filled.difference <- median.steps.per.days.filled - median.steps.per.days
# What is the impact of imputing missing data on the estimates of the total daily number of steps?
# print(paste("The impact of imputing the missing data on the estimates of the total daily number of steps is ",
#             mean.steps.filled.difference, "on the mean and ", format(median.steps.filled.difference, digits=2), "on the median"))
```

The total number of missing values in the dataset is **`r nb.missing.rows`**

For filling in all of the missing values in the dataset, the mean for that 5-minute interval across all days was used.

Mean of the total number of steps per day (NA filled): **`r format(mean.steps.per.days.filled, digits=6)`**   
Median of the total number of steps per day (NA filled): **`r format(median.steps.per.days.filled, digits=6)`**

The impact of imputing the missing data on the estimates of the total daily number of steps is 
**`r mean.steps.filled.difference`** , on the mean and **`r format(median.steps.filled.difference, digits=2)`** on the median.

## Are there differences in activity patterns between weekdays and weekends?
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Are there differences in activity patterns between weekdays and weekends?
# For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.
# 1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
data.filled.na$day <- weekdays(data.filled.na$date, abbreviate = TRUE)
data.filled.na$day[data.filled.na$day == "Sat"|data.filled.na$day == "Sun"] <- "weekend"
data.filled.na$day[data.filled.na$day != "weekend"] <- "weekday"


data.filled.na.avg <- data.filled.na %>% 
                      group_by(interval, day) %>% 
                      summarise(avg_steps=mean(steps))

# 2.Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and 
# the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 
# See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.
library("lattice")
xyplot(avg_steps ~ interval | day,
       data = data.filled.na.avg,
       type = "l", 
       layout = c(1,2),
       main="Average number of steps per interval for weekdays and weekend",
       xlab="Interval",
       ylab="Average number of steps")
```